Reconstructing Propositional Proofs in Type Theory

### 2017-10-26 paper/sections/introduction.tex
[ paper ] added some corrections.

An automatic theorem provers (henceforth \ATP) is a tool that intends to prove
conjectures based on assumptions and hypothesis in a logic specification.
Verifying proofs generated by \ATPs are non-trivial tasks.
In the last decades, \ATPs are fast becoming a key instrument in
different disciplines and real applications (\eg, verifying railway interlocking systems in~\cite{Kanso2012}).
Therefore, researchers from academy and industry have shown an increased interest to prove the validity of their deduction algorithms.

See for instance bugs found in ATPs in~\cite{Keller2013,Bohme2011,Fleury2014}.
For that reason, in order to give confidence to the user of these systems,
many ATPs have started to include in their outputs, a witness of their
conclusions. However, existing research recognizes that in many cases these proofs encode non-trivial reasoning
hard to reconstruct and therefore hard to verify. The proof-reconstruction
address this problem.
It becomes in mostly cases a reverse engineering to justify the prover reasoning.
That happen when the reconstruction is made by another and not by
the developers of the \ATP.

@@

One approach to address the proof-reconstruction
is integrating the \ATP, the \emph{source} system, with a proof-assistant, the
\emph{target} system.
The target system becomes in the proof checker. These proof-assistants allow
us to define the formal language used in the proofs, \ie, operators, logic
variables, axioms, and theorems. A proof-reconstruction tool provides such an
integration, translating the replied generated by the prover into the formalism of the proof-assistant.

>>>>

An automatic theorem prover (henceforth \ATP) is a tool that intends to prove
conjectures based on assumptions and hypothesis in a logic specification.
Verifying proofs generated by \ATPs are non-trivial tasks.
In the last decades, \ATPs are fast becoming a key instrument in
different disciplines and real applications (\eg, verifying railway
interlocking systems in~\cite{Kanso2012}).
Therefore, researchers from academy and industry have shown an increased
interest to prove the validity of their deduction algorithms since
some bugs have been found in these
tools~\cite{Keller2013,Bohme2011,Fleury2014}.

For that reason, in order to give confidence to the \ATP users,
many of these system have started to include in their outputs the derivations, witnesses, of their conclusions.
However, existing research recognizes that in many cases these proofs encode
non-trivial reasoning
hard to reconstruct and therefore hard to verify~\cite{paulson2007source,Keller2013}.

The proof-reconstruction address this problem. It becomes in mostly cases a reverse engineering to justify the prover reasoning. That happen when the reconstruction is made by another and not by the developers of the \ATP.

@@

One approach to address the proof-reconstruction problem is integrating the
\ATP, the \emph{source} system, with a proof-assistant, the \emph{target}
system. The target system is the proof checker in charge to  verify the prover reasoning.
These proof-assistants allow us to define the formal
language used in the proofs, \ie, operators, logic variables, axioms, and
theorems. A proof-reconstruction tool provides such an integration,
translating the replied generated by the prover into the formalism of the
proof-assistant.

### 2017-10-26 paper/sections/limitations-type-theory.tex
[ paper ] added some corrections.

\begin{notation}
We write the type judgments as $m : \abbre{M}$ to denote
that $m$ has $\abbre{M}$ type.
Types are written using small caps.
We use \Prop for the type of propositions and the dependent type, $Γ ⊢ φ$ for
theorems where $Γ : \List~\Prop$ and $φ : \Prop$.
$\abbre{List}$ type is the usual inductive type for lists.
For syntactical equivalence, we write the equality as $φ ≡ ψ$
for $φ,~ψ : \Prop$. For this equality, we assume the
reflexivity, symmetry and transitivity properties.
\end{notation}

@@

Algorithms based on general recursions can not be translated
directly in type theory since it is not a guarantee they terminate. For that
reason, we follow the technique described in~\cite{Bertot2004} to avoid
termination problems by modifying the recursive functions to be
\emph{structural recursive}.
% (see \cite{Coquand1992,Abel2002,Bove2005} for more details).

@@

General recursive functions can be translated into structurally recursive
functions by using for instance, the \emph{Bounded Recursion} technique
presented in~\cite{Bertot2004}. For a yet another methods, we refer
the reader to \cite{Coquand1992,Abel2002,Bove2005}.
The bounded technique defines a new structural recursive function based on
the general recursive function by adding an argument to the function.
The new argument is the \emph{bound}, a natural number given by the
function complexity. In other words, the added argument will represent
the number of times the function needs to call itself to get the
expected outcome~\cite{Bertot2004}.

@@

In~\eqref{eq:uh-definition}, the first two equation
call on arguments no structurally smaller.
For instance, on the first call, $φ ≡ (φ₁ ∧ φ₂) ⇒ φ₃$, the recursion
calls on the  $φ₁ ⇒ (φ₂ ⇒ φ₃)$ argument that is not a subformula of φ.
Therefore, one way to translate $\fuh_{0}$ to type
theory is to define a new function using a bounded recursion.

@@

We bounded the recursion calls in~\eqref{eq:uh-structured} with \Nat type argument. This natural number represents the complexity measure of $\fuh_{0}$
function.

Recall the \emph{complexity measure} defined in \cite{Agudelo-Agudelo2017}
and presented in \cite{VanDalen1994} as \emph{the rank for a
proposition}, is a function $\Prop \to \Nat$ that assigns a natural
number for a proposition $φ$ like the following $\frank$ function.
The symbols $c_{¬}$ and $c_{\square}$ are non-negative integer
numbers called the complexity of the respective connective.

>>>>

\begin{notation}
Types are written using small caps.
We write the type judgments as $m : \abbre{M}$ to denote
that $m$ has $\abbre{M}$ type.
We use \Prop for the type of propositions and the dependent type, $Γ ⊢ φ$ for
theorems where $Γ : \List~\Prop$ and $φ : \Prop$.
$\abbre{List}$ type is the usual inductive type for lists.
For syntactical equivalence, we write the equality as $φ ≡ ψ$
for $φ,~ψ : \Prop$. For this equality, we assume the
reflexivity, symmetry and transitivity properties.
\end{notation}

@@

Algorithms based on general recursion can not be translated
directly in type theory since it is not a guarantee they terminate. For that
reason, we follow the technique described in~\cite{Bertot2004} to avoid
termination problems by modifying the recursive functions to be
\emph{structural recursive}.
% (see \cite{Coquand1992,Abel2002,Bove2005} for more details).

@@

General recursive functions can be rewrote into structurally recursive
functions by using for instance, the \emph{bounded recursion} technique.
For a yet another methods, we refer
the reader to \cite{Coquand1992,Abel2002,Bove2005}.
The bounded technique defines a new structural recursive function based on
the general recursive function by adding an argument to the function.
The new argument is the \emph{bound}, a natural number given by the
function complexity. In other words, the added argument will represent
the number of times the function needs to call itself to get the
expected outcome~\cite{Bertot2004}.

@@

In~\eqref{eq:uh-definition}, the first two equations
call on arguments no structurally smaller.
For instance, on the first call, $φ ≡ (φ₁ ∧ φ₂) ⇒ φ₃$, the recursion
calls on the  $φ₁ ⇒ (φ₂ ⇒ φ₃)$ argument that is not a subformula of φ.
Therefore, one way to translate $\fuh_{0}$ to type
theory is to define a new function using a bounded recursion.

@@

We bounded the recursion calls in~\eqref{eq:uh-structured} with a \Nat argument. This natural number represents the complexity measure of $\fuh_{0}$
function.

The \emph{complexity measure} defined in \cite{Agudelo-Agudelo2017}
and presented in \cite{VanDalen1994} as \emph{the rank for a
proposition}, is a function $\Prop \to \Nat$ that assigns a natural
number to a formula like we do with the following $\frank$ function.
The symbols $c_{¬}$ and $c_{\square}$ are non-negative integer
numbers called the complexity of the respective connective.

### 2017-10-26 paper/sections/metis-language-and-proofs.tex
[ paper ] added some corrections.

\begin{figure}
\begin{equation*}
  \begin{bprooftree}
    \AxiomC{}
    \RightLabel{axiom}
    \UnaryInfC{$Γ ⊢ φ$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{}
    \RightLabel{assume $L$}
    \UnaryInfC{$Γ ⊢ φ ∨ ¬ φ$}
  \end{bprooftree}
  \end{equation*}
  \vskip2mm
  \begin{equation*}
  \begin{bprooftree}
    \AxiomC{$Γ ⊢ l ∨ φ$}
    \AxiomC{$Γ ⊢ ¬ l ∨ ψ$}
    \RightLabel{resolve $l$}
    \BinaryInfC{$Γ ⊢ φ ∨ ψ$}
  \end{bprooftree}
\end{equation*}
\caption{CPL inference rules of the \texttt{Metis} prover.}
\label{fig:metis-inferences}
\end{figure}

@@

      Strip a goal into subgoals
      &\texttt{strip}
      &\ref{thm:strip}
      \\

      Takes a formula from a conjunction
      &\texttt{conjunct}
      &\ref{thm:conjunct}
      \\

      A general form of the resolution theorem
      &\texttt{resolve}
      &\ref{thm:resolve}
      \\

      Performs clausification
      &\texttt{clausify}
      &\ref{thm:clausify}
      \\

      Normalization of the formula
      &\texttt{canonicalize}
      &\ref{thm:canonicalize}
      \\

      Simplify definitions and theorems
      &\texttt{simplify}
      &\ref{thm:simplify}
      \\[1ex]
    \hline
    \end{tabular}}
  \end{center}
\end{table}

>>>>

\begin{figure}
\begin{equation*}
  \begin{bprooftree}
    \AxiomC{}
    \RightLabel{axiom}
    \UnaryInfC{$Γ ⊢ φ$}
  \end{bprooftree}
  \qquad
  \begin{bprooftree}
    \AxiomC{}
    \RightLabel{assume $L$}
    \UnaryInfC{$Γ ⊢ φ ∨ ¬ φ$}
  \end{bprooftree}
  \end{equation*}
  \vskip2mm
  \begin{equation*}
  \begin{bprooftree}
    \AxiomC{$Γ ⊢ l ∨ φ$}
    \AxiomC{$Γ ⊢ ¬ l ∨ ψ$}
    \RightLabel{resolve $l$}
    \BinaryInfC{$Γ ⊢ φ ∨ ψ$}
  \end{bprooftree}
\end{equation*}
\caption{Inference rules of the \texttt{Metis} prover.}
\label{fig:metis-inferences}
\end{figure}

@@

      \texttt{strip}
      &Strip a goal into subgoals
      &\ref{thm:strip}
      \\

      \texttt{conjunct}
      &Takes a formula from a conjunction
      &\ref{thm:conjunct}
      \\

      \texttt{resolve}
      &A general form of the resolution theorem
      &\ref{thm:resolve}
      \\

      \texttt{clausify}
      &Performs clausification
      &\ref{thm:clausify}
      \\

      \texttt{canonicalize}
      &Normalization of the formula
      &\ref{thm:canonicalize}
      \\

      \texttt{simplify}
      &Simplify definitions and theorems
      &\ref{thm:simplify}
      \\[1ex]
    \hline
    \end{tabular}}
  \end{center}
\end{table}

### 2017-10-26 paper/sections/related-work.tex
[ paper ] added some corrections.

In type theory, for \Agda and  for the same propositional fragmented used in this paper,
we found the proof-reconstruction proposed by \citeauthor{Kanso2012}
in~\cite{Kanso2012,kanso2016light}. The author present an approach similar to
the reasoning presented in Section~\ref{ssec:workflow}.
Nonetheless, his approach includes semantics reasoning. We have avoided the
use of propositions meanings towards a future work to support other logics but
also contrast the usage of a deep embedding of the logic with a shallow
embedding.

>>>>

For type theory using \Agda and for the same propositional fragmented used in this paper, we found the proof-reconstruction proposed by \citeauthor{Kanso2012} in~\cite{Kanso2012,kanso2016light}.
The author present an approach similar to
the reasoning presented in Section~\ref{ssec:workflow}.
Nonetheless, his approach includes semantics reasoning. We have avoided the
use of propositions meanings towards a future work to support other logics but
also contrast the usage of a deep embedding of the logic with a shallow
embedding.
